What We Eval in the Shadows

### 2021-04-11 analysis/environments.Rmd
Environment

overview_table(
  r("default envir sites", nb_default_sites),
  r("defaut envir sites percent", ratio(nb_default_sites, nb_call_sites)),
  r("non default envir sites", nb_explicit),
  r("non defaut envir sites percent", ratio(nb_explicit, nb_call_sites))
)
```

@@

```{r envir_absolute}
eval_calls_absolute <- eval_calls_explicit %>% filter(envir_type == "environment")

@@

# TODO: handle scientific notation for caller-1e+00-

env_calls <- eval_calls_absolute %>% ungroup() %>%
  extract(environment_class, regex = "((?:new\\+)*)caller-([[:digit:]]+)-(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
  mutate(new_env = na_if(new_env, ""), specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) %>%
  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env))
```

@@

@@

`caller-n` wins over any specific env. We add `+` if there are any `new`.

TODO: if there is both new , a caller-n, a specific class, rather ignore the caller-n
```{r}
env_classes <- env_calls %>%
  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(hierarchy), specific_env, as.character(hierarchy)), if_else(is.na(new_env), "", "+"))))

env_classes %>% count(env_class, wt = duplicates, sort = TRUE) %>%
  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc)) %>%
  datatable()

env_classes %>% group_by(env_class) %>% summarise(n = n_distinct(src_ref)) %>%
  arrange(desc(n)) %>%
  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc)) %>%
  datatable()
```

>>>>

overview_table(
  r("default envir sites", nb_default_sites),
  r("defaut envir sites percent", ratio(nb_default_sites, nb_call_sites)),
  r("non default envir sites", nb_explicit_sites),
  r("non defaut envir sites percent", ratio(nb_explicit_sites, nb_call_sites))
)
```

@@

```{r envir_absolute}
eval_calls_absolute <- eval_calls_explicit %>% filter(envir_type %in% c("environment", "S4"))

@@

env_calls <- E_envir %>% filter(envir_type %in% c("environment", "S4")) %>%
  extract(environment_class, regex = "((?:new\\+)*)caller-([[:digit:]]+)-(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
  mutate(new_env = na_if(new_env, ""), specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) %>%
  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env)) %>% 
  mutate(hierarchy = hierarchy - 1) # Fix. Remove with the next run
```

@@

Glabal can happen by chance, because `parent.frame()`goes up too much and reaches the global environment. SO when is it explicitly passed?

```{r}
explicit_global <- E_envir %>% filter(envir %in% c("globalenv()", ".GlobalEnv"))

nb_explicit_global_sites <- explicit_global %>% pull(src_ref) %>% n_distinct()

overview_table(
  r("nb explicit global sites", nb_explicit_global_sites),
  r("nb explicit global site percent", ratio(nb_explicit_global_sites, nb_call_sites))
)
```

@@

A specific env wins over `caller-n`. We add `+` if there are any `new`.

```{r}
env_classes <- env_calls %>%
  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(specific_env), as.character(hierarchy), specific_env), if_else(is.na(new_env), "", "+"))))

not_env_calls <- E_envir %>% anti_join(env_classes) %>% 
  mutate(env_class = case_when(
    envir_type == "list" ~ "list",
    envir_type == "integer" ~ "int",
    envir_type == "NULL" ~ "NULL",
    TRUE ~ "other"
  ))

env_classes <- bind_rows(env_classes, not_env_calls)

env_classes_calls <- env_classes %>% count(env_class, wt = duplicates, sort = TRUE) %>%
  mutate(pc = ratio(n, nb_eval_calls), cpc = cumsum(pc))

env_classes_sites <- env_classes %>% group_by(env_class) %>% summarise(n = n_distinct(src_ref)) %>%
  arrange(desc(n)) %>%
  mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc)) 
```

```{r}
env_classes_calls %>% datatable()

env_classes_sites %>% datatable()

nrows <- 8
r_vec("env class names", nrows, env_classes_sites$env_class)
r_vec("env class sites", nrows, env_classes_sites$n)
r_vec("env class site percent", nrows, env_classes_sites$pc)
```

### 2021-04-11 paper/main.tex
Environment

\c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
list or data frame and creates bindings for them in a new environment.

@@

We looked at the environments passed to all the \eval calls in our corpus.
Table~\ref{tab:environments} summarizes the results. A numeric environment class
\c{n} denotes the environment of the $n$-th call stack frame from the current
function. \c{global} denotes the top-level environment and \c{list} denotes a
list passed for evaluation of formulas. Environment classes of the form $n+$
denote the $n$-th environment extended with a new environment. The new
environment provides a limited form of sandboxing. All assignments using the
$\leftarrow$ function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment using
the $\leftarrow$ or \c{assign} functions; but, that happens rarely.

% This table is old and incorrect
% Working on updating it
\begin{table}[htbp]{ \centering
\begin{tabular}{|c|c|}\hline
 Environment Class & Eval Calls\\\hline
 \PackagesEnvClassA &  \PackagesProportionA\\
 \PackagesEnvClassB &  \PackagesProportionB\\
 \PackagesEnvClassC &  \PackagesProportionC\\
 \PackagesEnvClassD &  \PackagesProportionD\\
\PackagesEnvClassE &  \PackagesProportionE\\\hline
\end{tabular}} \label{tab:environments}
\caption{Environments in terms of \eval calls}
\end{table}

@@

Categorize side-effects between env passed to eval and other envs:
\begin{itemize}
\item ephemeral -- read to envs that are created during \eval  XXX033
\item local -- reads to env of the function in which the call to \eval
  lexically occurred; XXX034
\item parent -- reads to parent envs of the local one. XXX035
\item call-stack -- reads to envs that are on the call stack but not
  caller. XXX035.
\end{itemize}

>>>>

It can also be \c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
list or data frame and creates bindings for them in a new environment.

We looked at the environments passed to all the \eval calls in our corpus.
Table~\ref{tab:environments} summarizes the results. A numeric environment class
\c{n} denotes the environment of the $n$-th call stack frame from the current
function: \c{0} is the stack of the current function, \c{1} is the parent frame. \c{global} denotes the top-level environment and \c{list} denotes a
list passed for evaluation of formulas. Environment classes of the form $n+$
denote the $n$-th environment extended with a new environment. The new
environment provides a limited form of sandboxing. All assignments using the
$\leftarrow$ function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment using
the $\leftarrow$ or \c{assign} functions; but, that happens rarely.

% Updated with the last run
% I made some changes and we need another run to get the right results
% The difference should be that there will much more package environments
% the loop category should disappear from the top results also.
\begin{table}[htbp]{ \centering
        \begin{tabular}{|c|c|}\hline
            Environment Class & Sites\\\hline
            \packageEnvClassNamesa &  \packageEnvClassSitePercenta\\
            \packageEnvClassNamesb &  \packageEnvClassSitePercentb\\
            \packageEnvClassNamesc &  \packageEnvClassSitePercentc\\
            \packageEnvClassNamesd &  \packageEnvClassSitePercentd\\
            \packageEnvClassNamese &  \packageEnvClassSitePercente\\
            \packageEnvClassNamesf &  \packageEnvClassSitePercentf\\
            \packageEnvClassNamesg &  \packageEnvClassSitePercentg\\
            \packageEnvClassNamesh &  \packageEnvClassSitePercenth\\
            \hline
    \end{tabular}} \label{tab:environments}
    \caption{Environments in terms of sites. One site can be passed different kinds of environments.}
\end{table}

Most of the calls happen in the environment surrounding \eval. This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument. A lot of sites also use \c{parent.frame()} to access their parent frame. It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level. Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.

@@

@@

\todo{That should be moved to the side effect section and updated with the last data.}
Categorize side-effects between env passed to eval and other envs:
\begin{itemize}
\item ephemeral -- read to envs that are created during \eval  XXX033
\item local -- reads to env of the function in which the call to \eval
  lexically occurred; XXX034
\item parent -- reads to parent envs of the local one. XXX035
\item call-stack -- reads to envs that are on the call stack but not
  caller. XXX035.
\end{itemize}
